---
title: Testing of the mvgam package
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    css: style.css
    toc: true
    toc-float: true
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
---

```{r chunks, results='markdown', eval=TRUE, echo = FALSE}
assignInNamespace('.sep.label',  "^\\ *(#|--)+\\s*(@knitr|----+)(.*?)-*\\s*$", ns='knitr')

knitr::opts_chunk$set(tidy = "styler")

knitr::read_chunk("../R/analysis.R") 
```


# Preparations

- Load required packages
```{r}
#| label: load_libraries
#| eval: true
#| message: false
#| warnings: false
#| output: false
#| cache: false
```

- Load other functions
```{r}
#| label: load_functions
#| eval: true
#| message: false
#| warnings: false
#| output: false
#| cache: false
```

# Create fabricated data

Lets create some simple motivating data. These data will represent a
simplified version of the kinds of data that are very common in
ecological monitoring programs.  In this case, the design comprises:

- **six** large scale locations (for example, a selection of
  individual coral **reefs** within a portion of the Great Barrier
  Reef).
- within each reef, there are **three fixed sites** (distinct,
  permanently allocated representative areas within each reef)
- each site is surveyed annually for 10 years (**time**) 
- the amount (percentage cover) of three **species** (for example
  coral taxa) are monitored from each site, each year
- to keep  things simple for this example, we will assume that coral
  cover is generated from a Gaussian distribution
- this design is represented in @fig-visualise_design_1 and
  @fig-visualise_design_2 below

Notes:

- typically, monitoring designs comprise additional spatial scales.
  For example, there are often multiple transects within each site.
- individual coral cover observations are usually derived from
  identifying individual corals under a grid of points on photographs
  taken along the transects. As such, coral cover usually modelled
  against a binomial distribution (points of the focal taxa, total
  photo points)
- there are usually many more than three species identified
- long-term monitoring on the Great  Barrier Reef extends back to 1985
  (hence approaching 40 years of continual monitoring)

::: {.callout-note}
The data generated in this section are saved (as a tibble) in binary
format towards the end of the section and read back in the proceeding
section. As a result, the current section on creating the fabricated
data can be skipped.
:::



::: {.panel-tabset}

## Define parameters

```{r}
#| label: define parameters
#| results: markup
#| eval: true
#| echo: true
#| cache: false
```
## Define reefs

```{r}
#| label: define reefs
#| results: markup
#| eval: true
#| echo: true
#| cache: false
```

## Define sites

```{r}
#| label: define sites
#| results: markup
#| eval: true
#| echo: true
#| cache: false
```

## Define species

```{r}
#| label: define species
#| results: markup
#| eval: true
#| echo: true
#| cache: false
```

## Define times

```{r}
#| label: define times
#| results: markup
#| eval: true
#| echo: true
#| cache: false
```
## Define splines

```{r}
#| label: define splines
#| results: markup
#| eval: true
#| echo: true
#| cache: false
```
## Define Gaussian response

```{r}
#| label: define Gaussian response
#| results: markup
#| eval: true
#| echo: true
#| cache: false
```
:::

# Load the data

```{r}
#| label: load_data
#| results: markup
#| eval: true
#| echo: true
#| cache: false
```



# Visualise design

```{r}
#| label: fig-visualise_design_1
#| results: markup
#| eval: true
#| echo: true
#| cache: false
#| fig-width: 20
#| fig-height: 8
#| fig-cap: design of fabricated data illustrating the spatio-temporal hierarchy.  Note, species have been omitted to save space.
```

```{r}
#| label: fig-visualise_design_2
#| results: markup
#| eval: true
#| echo: true
#| cache: false
#| fig-width: 20
#| fig-height: 8
#| fig-cap: design of fabricated data illustrating further details of the spatio-temporal hierarchy. Note, the full species and temporal design is only shown for a single site so as to save space.
```

# Exploratory data analysis

```{r}
#| label: fig-eda_1
#| results: markup
#| eval: true
#| echo: true
#| cache: false
#| fig-width: 12
#| fig-height: 10
#| fig-cap: Temporal trends in three coral species across multiple sites and reefs
```

```{r}
#| label: fig-eda_2
#| results: markup
#| eval: true
#| echo: true
#| cache: false
#| fig-width: 12
#| fig-height: 10
#| fig-cap: Simplified temporal trends in three coral species across multiple sites and reefs
```

# Data preparations

In order to access the convenience of the `mvgam` package, it is
necessary to prepare the data a little:

- the multivariate components need to be in a factor called `series`
- the "sites" need to be uniquely identified within each "reef" (in
  this case they already were)
- the temporal component needs to be a numeric variable representing
  the time increments

```{r}
#| label: data preparation
#| results: markup
#| eval: true
#| echo: true
#| cache: false
```

# Data analysis

We seem to have hit a snag on the first hurdle - exploring the default
priors (although this is likely going to also be an issue downstream).

The issue is that with this hierarchical design, there are multiple
instances of each time point (one for each Reef/Site/Species).
Internally, this is triggering an error designed to highlight missing
observations or timepoints.

This issue is highlighted via a number of explorations:

1. attempting to gain priors for a null model from the full design -
   fails
2. attempting to gain priors for a null model from a design aggregated
   to a single value per timepoint - succeeds
3. attempting to gain priors for a null model from a design aggregated
   to the Reef/Timepoint level (hence one timepoint per six reefs) - fails
4. it also fails when we attempt to gain priors for structures that
   include Reef as a random effect in either the observation or
   process model formulae (on data aggregated to the Reef/Timepoint
   level)

::: {.panel-tabset}

## Exploration 1

```{r}
#| label: analysis exploration 1
#| results: markup
#| eval: true
#| error: true
#| echo: true
#| cache: false
```
## Exploration 2
```{r}
#| label: analysis exploration 2
#| results: markup
#| eval: true
#| error: true
#| echo: true
#| cache: false
```
## Exploration 3
```{r}
#| label: analysis exploration 3
#| results: markup
#| eval: true
#| error: true
#| echo: true
#| cache: false
```
:::
